Discrepancy, Coresets, and Sketches in Machine Learning

### 2019-01-28 komlos.tex

\begin{definition}[\bf Class Discrepancy] \label{cd1}
Let $A \subset \R^m$ and $\sigma \in \{-1,1\}^m$ the class discrepancy of $A$ is 
\[
K_m(A)  = \min_\sigma \max_{a\in A} \left| \frac{1}{m} \sum_{i=1}^{m}  \sigma_i  a_i\right| 
\]
\end{definition}
%
\begin{definition}[\bf Class Discrepancy]\label{cd2}
Let $f:\X,\Q\rightarrow\R$ and $\sigma \in \{-1,1\}^m$. 
The class discrepancy of $f$ w.r.t.\ $\{x_1,\ldots,x_m\} \subset \X$ is 
\[
K_m(f) =  \min_{\sigma} \max_{q \in \Q}  \left|\frac{1}{m} \sum_{i=1}^{m}\sigma_i f(x_i,q)\right|
\]
\end{definition}
%
\begin{definition}[\bf Class Discrepancy]\label{cd3}
Let $\F$ be a family of functions $f:\X\rightarrow\R$ and $\sigma \in \{-1,1\}^m$. 
The class discrepancy of $\F$ w.r.t.\ $\{x_1,\ldots,x_m\} \subset \X$ is 
\[
K_m(\F) =  \min_{\sigma} \max_{f \in \F}  \left| \frac{1}{m} \sum_{i=1}^{m}\sigma_i f(x_i)\right|
\]
\end{definition}
The class discrepancy of $f$ or $\F$ without a reference a set $\{x_1,\ldots,x_m\}$ is the upper bound on any such set.
Throughout the manuscript we assume a one to one mapping between $\F$ and $\Q$. 
Specifically, $f \in \F$ if there exists $q \in \Q$ such that $f(x) = f(x,q)$.
In the context of machine learning, one should think about $f(x,q)$ as the loss associated with example $x$ and model parameters $q$.
Definitions~\ref{cd2} and \ref{cd3} are therefore equivalent. 
For the equivalence of Definition~\ref{cd1}, the set $A$ should be thought of as the set of possible induces losses. 
Namely, $a\in A$ if there is a model $q$ such $a_i = f(x_i,q)$.
The three equivalent definitions between the definitions is therefore straight forward.

@@

To understand our motivation, consider the following informal explanation of the Radamacher Complexity applied to ML problems. 
In PAC learning there exists a set of examples (with labels).
We aim to find a regressor/classifier from a given family that suffers the least loss on the set. 
Having a low Radamacher complexity means that we can optimize over a sample of roughly half the examples at random (each w.p.\ $1/2$).
Low Rademacher complexity guaranties that, in expectation, twice the loss on the sample is roughly the same as the loss on the entire set.
This translates to a generalization bound. 
In other words, the Radamacher complexity gives a guarantee for the loss of coresets chosen uniformly at random.

Coming back to discrepancy. Having the ability to choose the signs arbitrarily lets us choose an advantageous subset of examples.
We can algorithmically choose to minimize the induced error and guaranty to have (roughly) the same performance on the entire set. 
This set is, in fact, a coreset. The class discrepancy of a problem helps us determine what coreset sized are possible. 
We will show in the following sections several examples for which a coreset can be significantly smaller than the random sample while maintaining the same guarantees. This will be done by showing that for a wide range of interesting problems in machine learning $K_m(\F) = o(R_m(\F))$.
This intuition is restated more explicitly in the next section.

@@

\noindent The following facts are true for the common cases where $K_m = O(c/m)$ or $K_m = O(c/\sqrt{m})$.
\begin{fact}
Any function $f$ with Class Discrepancy $K_m(f) = O(c/m)$ has Coreset Complexity of $f$ $O(c/\eps)$.
The constant $c$ does not depend on $m$.
\end{fact}
\begin{fact}
Any function $f$ with Class Discrepancy $K_m(f) = O(c/\sqrt{m})$ has Coreset Complexity $O(c^2/\eps^2)$.
The constant $c$ does not depend on $m$.
\end{fact}
\begin{proof}
\noindent

Consider the signed sum of error function $E(q) = \sum_{i=1}^{m} \sigma_i f(x_i,q)$ where $\sigma_i \in \{-1,1\}$.
Now consider $\tilde F_{+}(q) = F(q) + E(q)   = \sum_{i | \sigma_i=1} 2 f(x_i,q)$ and similarly $\tilde F_{-}(q) = F(q) - E(q)  =  \sum_{i | \sigma_i=-1} 2 f(x_i,q)$. We have that both $\tilde F_{+}(q)$ and $\tilde F_{-}(q)$ are approximations for $F(q)$ obtained by coresets of item weights of $2$ and an error at most $|\tilde F_{\pm}(q)- F(q)| =  |E(q)|$. The above is true for any choice of signs $\sigma$, specifically, for those minimizing $\max_q | E(q)|$.
By Definition we can select signs such that $|E(q)| \le m K_m(f)$.

@@

\section{Related Work}
\begin{itemize}
\item coresets - motivation, some papers, streaming algorithms for coresets

\item Radamacher complexity \cite{Bartlett:2003:RGC:944919.944944} is a known method to estimate the generalization ability of an empirical risk minimizer (ERM) of a family of function over a finite set of i.i.d sampled points. A carefully selected set of points can potentially obtain a better generalization bound, and that is the usefulness of coresets in the context of machine learning. We are not aware of any analog to Radamacher complexity that aims to measure the generalization ability of a coreset of a fixed finite size. There are papers such as \cite{tolochinsky2018coresets} and references within that tie the coreset to the VC dimension of the family function, and the average sensitivity of the dataset. However, these relationships come up more as tools for constructing a coreset rather than a single complexity measure aimed to characterize some generalization ability.
%
% There is literature about the complexity of active learning for the supervised setting but that is out of scope as we deal with the setting where all labels are available, or the problem is unsupervised. For the complexity of coresets, the total sensitivity of the dataset and the VC dimension of the function class is often used as tools for construction \zk{cite something?}. However, to the best of our knowledge there is no clear characterization of the size of coresets based on some complexity measure of the dataset and function class.

%
%\item Radamacher complexity \cite{Bartlett:2003:RGC:944919.944944}, There is literature about the complexity of active learning for the supervised setting but that is out of scope as we deal with the setting where all labels are available, or the problem is unsupervised. For the complexity of coresets, the total sensitivity of the dataset and the VC dimension of the function class is often used as tools for construction \zk{cite something?}. However, to the best of our knowledge there is no clear characterization of the size of coresets based on some complexity measure of the dataset and function class.

\item Kernel density estimation is a popular tool in data analysis aimed to estimate a continuous distribution with a finite set of points. Among other applications, this tool is used for outlier detection \cite{schubert2014generalized}, regression \cite{fan2018local}, and clustering \cite{rinaldo2010generalized}. A thorough survey could be found in \cite{silverman2018density}. Given a set of $n$ data points $x_1,\ldots,x_n$ and a query $q$, the density estimate of $q$ is the average $\frac{1}{n} \sum_i K(x_i,q)$ of the kernel evaluations with the data points. The task of finding a coreset for kernel density estimates consists of obtaining a set of points $z_1,\ldots,z_m$ with weights $w_1,\ldots,w_m$ for some $m \ll n$ such that for any query $q$ it holds that 
$\frac{1}{n} \sum_i K(x_i,q) = \frac{1}{m} \sum_i w_i K(z_i,q) \pm \eps$. For this task, the state-of-the-art is given by \cite{DBLP:journals/corr/abs-1802-01751}, achieving $m=\sqrt{d\log(1/\eps)}/\eps$ where $d$ is the dimension of the original data points. Their result holds for Lipchitz bounded positive semi-definite kernels. The result is constructive though it is polynomial rather than (quasi-)linear in the data size. The authors give an almost matching lower bound of $\sqrt{d}/\eps$ and pose an open question for closing the gap between the bounds.

@@

\paragraph{Theorem~\ref{thm:streaming11}}
For any function family $\F$ with a corresponding discrepancy $K_m = O(c/m)$ there exists an fully-mergeable streaming coreset deterministic algorithm of size $O\left(c\log^2(\eps n/c)/\eps\right)$ whose error is at most $\eps n$.
\begin{proof}
Consider feeding the output of the first compactor into a second one etc. Specifically, we start with a single compactor and open a second once it produced any output, then open a third compactor once the second produced output, etc.
Number the compactors $0,\ldots,H$. The weight of items given to compactors $h$ have weight $w_h = 2^h$. The length of the input stream seen by compactor is $n_h \le n/2^h$.
Each compactor contributes at most $w_h n_h K_m \le n K_m$ error. Moreover since $n_h \le n/2^h$ \todo{and compactor number $H-1$ observed at least $m$ items we know that $H \le \floor{log_2(n/m)}+1$. TODO: Not sure this is going in the right direction...}
The total error is therefore $H n K_m \le O(log(n/m) n K_m)$. 
Setting $m \ge m_0 = O(c\log(\eps n/c)/\eps)$ and replacing $K_m = c/m$ we get that the error is at most $O(log(n/m) n K_m) \le \eps n$ for appropriate constants in the $O()$ term of $m_0$, as required. 
Since we have $H = O(log(\eps n/c))$ compactors the overall space complexity is $O(c\log^2(\eps n/c)/\eps)$. 
\end{proof}

>>>>

\begin{definition}[\bf Class Discrepancy] \label{cd1}
Let $A \subset \R^m$ and $\sigma \in \{-1,1\}^m$ the class discrepancy of $A$ is 
\[
K_m(A)  = \min_\sigma \max_{a\in A} \left| \frac{1}{m} \sum_{i=1}^{m}  \sigma_i  a_i\right| 
\]
\end{definition}
%
\begin{definition}[\bf Class Discrepancy]\label{cd2}
Let $f:\X,\Q\rightarrow\R$ and $\sigma \in \{-1,1\}^m$. 
The class discrepancy of $f$ w.r.t.\ $\{x_1,\ldots,x_m\} \subset \X$ is 
\[
K_m(f) =  \min_{\sigma} \max_{q \in \Q}  \left|\frac{1}{m} \sum_{i=1}^{m}\sigma_i f(x_i,q)\right|
\]
\end{definition}
%
\begin{definition}[\bf Class Discrepancy]\label{cd3}
Let $\F$ be a family of functions $f:\X\rightarrow\R$ and $\sigma \in \{-1,1\}^m$. 
The class discrepancy of $\F$ w.r.t.\ $\{x_1,\ldots,x_m\} \subset \X$ is 
\[
K_m(\F) =  \min_{\sigma} \max_{f \in \F}  \left| \frac{1}{m} \sum_{i=1}^{m}\sigma_i f(x_i)\right|
\]
\end{definition}
The class discrepancy of $f$ or $\F$ without a reference a set $\{x_1,\ldots,x_m\}$ is the upper bound on any subset of $\X$ of size $m$.
Throughout the manuscript we assume a bijective mapping between $\F$ and $\Q$. 
Specifically, $f \in \F$ if there exists $q \in \Q$ such that $f(x) = f(x,q)$ and vice versa. 
In the context of machine learning, one should think about $f(x,q)$ as the loss associated with example $x$ and model parameters $q$.
Definitions~\ref{cd2} and \ref{cd3} are therefore equivalent. 
For the equivalence of Definition~\ref{cd1}, the set $A$ should be thought of as the set of possible induces losses. 
Namely, $a\in A$ if there is a model $q$ such $a_i = f(x_i,q)$.
The three equivalent definitions between the definitions is therefore straight forward.

@@

To understand our motivation, consider the following informal explanation of the Rademacher Complexity applied to ML problems. 
In PAC learning there exists a set of examples (often with labels).
We aim to find a regressor/classifier from a given family that suffers the least loss on the set. 
Having a low Rademacher complexity means that we can optimize over a sample of roughly half the examples at random (each w.p.\ $1/2$).
Low Rademacher complexity guaranties that, in expectation, twice the loss on the sample is roughly the same as the loss on the entire set.
This translates to a generalization bound. 
In other words, the Rademacher complexity gives a guarantee for the loss of coresets chosen uniformly at random.

Coming back to discrepancy. Having the ability to choose the signs arbitrarily lets us choose an advantageous subset of examples.
We can algorithmically choose to minimize the induced error and guaranty to have (roughly) the same performance on the entire set. 
This set is, in fact, a coreset. The class discrepancy of a problem helps us determine the obtainable coreset size. 
We will show in the following sections several examples for which a coreset can be significantly smaller than the random sample while maintaining the same guarantees. This will be done by showing that for a wide range of interesting problems in machine learning $K_m(\F) = o(R_m(\F))$.
This intuition is restated more explicitly in the next section.

@@

\noindent The following facts are true for the common cases where $K_m = O(c/m)$ or $K_m = O(c/\sqrt{m})$.
\begin{fact}
Any function $f$ with Class Discrepancy $K_m(f) = O(c/m)$ has Coreset Complexity of $f$ $O(c/\eps)$.
The constant $c$ does not depend on $m$.
\end{fact}
\begin{fact}
Any function $f$ with Class Discrepancy $K_m(f) = O(c/\sqrt{m})$ has Coreset Complexity $O(c^2/\eps^2)$.
The constant $c$ does not depend on $m$.
\end{fact}
\begin{proof}
\noindent  Consider the signed sum of error function $E(q) = \sum_{i=1}^{m} \sigma_i f(x_i,q)$ where $\sigma_i \in \{-1,1\}$.
Now consider $\tilde F_{+}(q) = F(q) + E(q)   = \sum_{i | \sigma_i=1} 2 f(x_i,q)$ and similarly $\tilde F_{-}(q) = F(q) - E(q)  =  \sum_{i | \sigma_i=-1} 2 f(x_i,q)$. We have that both $\tilde F_{+}(q)$ and $\tilde F_{-}(q)$ are approximations for $F(q)$ obtained by coresets of item weights of $2$ and an error at most $|\tilde F_{\pm}(q)- F(q)| =  |E(q)|$. The above is true for any choice of signs $\sigma$, specifically, for those minimizing $\max_q | E(q)|$.
By Definition we can select signs such that $|E(q)| \le m K_m(f)$.

@@

\section{Related Work}
\begin{itemize}
\item coresets - motivation, some papers, streaming algorithms for coresets.

\item Streaming coresets - MRL, etc. It is in fact known that for certain family types, a random sample of size $\log(1/\delta)/\eps^2$ is an $\eps$-coreset with probability $1-\delta$. For example, for families of VC dimension $v$, a sample of $(1/\eps^2)(v+\log(1/\delta))$ suffices \cite{talagrand1994sharper}.

\item Rademacher complexity \cite{Bartlett:2003:RGC:944919.944944} is a known method to estimate the generalization ability of an empirical risk minimizer (ERM) of a family of function over a finite set of i.i.d sampled points. A carefully selected set of points can potentially obtain a better generalization bound, and that is the usefulness of coresets in the context of machine learning. We are not aware of a complete analog to Rademacher complexity that aims to measure the generalization ability of a coreset of a fixed finite size. There are papers such as \cite{langberg2010universal, tolochinsky2018coresets} and references within that tie the coreset to the VC dimension of the family function, and the average sensitivity of the dataset. However, these relationships come up more as tools for constructing a coreset rather than a single complexity measure aimed to characterize some generalization ability. The connection between coresets and discrepancy have been known in the non-streaming setting. For example, \cite{phillips2009small} (Theorem 1.1) provide an analog to our Facts~\ref{fct:eps1},~\ref{fct:eps2}.
%
% There is literature about the complexity of active learning for the supervised setting but that is out of scope as we deal with the setting where all labels are available, or the problem is unsupervised. For the complexity of coresets, the total sensitivity of the dataset and the VC dimension of the function class is often used as tools for construction \zk{cite something?}. However, to the best of our knowledge there is no clear characterization of the size of coresets based on some complexity measure of the dataset and function class.

%
%\item Rademacher complexity \cite{Bartlett:2003:RGC:944919.944944}, There is literature about the complexity of active learning for the supervised setting but that is out of scope as we deal with the setting where all labels are available, or the problem is unsupervised. For the complexity of coresets, the total sensitivity of the dataset and the VC dimension of the function class is often used as tools for construction \zk{cite something?}. However, to the best of our knowledge there is no clear characterization of the size of coresets based on some complexity measure of the dataset and function class.

\item Kernel density estimation is a popular tool in data analysis aimed to estimate a continuous distribution with a finite set of points. Among other applications, this tool is used for outlier detection \cite{schubert2014generalized}, regression \cite{fan2018local}, and clustering \cite{rinaldo2010generalized}. A thorough survey could be found in \cite{silverman2018density}. Given a set of $n$ data points $x_1,\ldots,x_n$ and a query $q$, the density estimate of $q$ is the average $\frac{1}{n} \sum_i K(x_i,q)$ of the kernel evaluations with the data points. The task of finding a coreset for kernel density estimates consists of obtaining a set of points $z_1,\ldots,z_m$ with weights $w_1,\ldots,w_m$ for some $m \ll n$ such that for any query $q$ it holds that 
$\frac{1}{n} \sum_i K(x_i,q) = \frac{1}{m} \sum_i w_i K(z_i,q) \pm \eps$. For this task, the state-of-the-art is given by \cite{DBLP:journals/corr/abs-1802-01751}, achieving $m=\sqrt{d\log(1/\eps)}/\eps$ where $d$ is the dimension of the original data points. Their result holds for Lipchitz bounded positive semi-definite kernels. The result is constructive though it is polynomial rather than (quasi-)linear in the data size. The authors give an almost matching lower bound of $\sqrt{d}/\eps$ and pose an open question for closing the gap between the bounds. The authors provide a simpler algorithm based on the Frank-Wolf method that achieves a $1/\eps^2$ sized coreset. It is in fact known (see \cite{lopez2015towards}, Theorem 1) that a sample of $\log(1/\delta)/\eps^2$ points gives a coreset w.p. $1-\delta$ for some kernel types.

@@

\paragraph{Theorem~\ref{thm:streaming11}}
For any function family $\F$ with a corresponding discrepancy $K_m = O(c/m)$ there exists an fully-mergeable streaming coreset deterministic algorithm of size $O\left(c\log^2(\eps n/c)/\eps\right)$ whose error is at most $\eps n$.
\begin{proof}
Consider feeding the output of the first compactor into a second one etc. Specifically, we start with a single compactor and open a second once it produced any output, then open a third compactor once the second produced output, etc.
Number the compactors $0,\ldots,H$. The weight of items given to compactors $h$ have weight $w_h = 2^h$. The length of the input stream seen by compactor is $n_h \le n/2^h$.

Each compactor contributes at most $w_h n_h K_m \le n K_m$ error. Moreover since the $H-1$ layer had outputs, we must have $m \leq n_{H-1}$ and
$$  \log_2(m) \leq \log_2(n_{H-1}) \leq \log_2(n) - (H-1)$$
leading to a bound $H \le \floor{log_2(n/m)}+1$.
The total error is therefore $H n K_m \le O(log(n/m) n K_m)$. 
Setting $m \ge m_0 = O(c\log(\eps n/c)/\eps)$ and replacing $K_m = c/m$ we get that the error is at most $O(log(n/m) n K_m) \le \eps n$ for appropriate constants in the $O()$ term of $m_0$, as required. 
Since we have $H = O(log(\eps n/c))$ compactors the overall space complexity is $O(c\log^2(\eps n/c)/\eps)$. 
\end{proof}
